A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-10 paper/paper.tex
makeup for the paper finish

\section{Relation to other cognitive models}
\subsection{Predictive Coding Hypothesis}
The predictive coding and Bayesian brain framework 
is a frequently evoked idea related to the default mode function
\cite{bar2007}.
According to this view,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information. 
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
back projections probably outnumber by far the existing input projection in the brain
\cite{salin1995corticocortical}.
It can explains the constructive, generative nature of sensory perception
\cite{friston2010free} and
why motor action is intimately related to sensory expectations
\cite{wolpert1995internal}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back-projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is maintained by top-down modulation of sensorimotor
processing by context-specific a-priori information.
The generative model of how data arise in the
world hence be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the incoming sensory examples.

\begin{itemize}
  \item Both accounts are process theories backed up by plausible
  neurobiological evidence that
  view the brain as a \"statistical organ\"
  generalizing from the past to new experiences.
  \item Both expose a mechanisms of brain plasticity with increased
  adaption of neuronal wiring when faced by unprecedented environmental
  challenges.
  \item The hierarchical aspects from predictive coding
  is re-expressed in MDP in form of
  nested prediction of probable upcoming actions and rewards.
  \item Both model the consequences of action. In RL, the horizon of that
  look into the future is manifest in the $\gamma$ parameter
  in the Bellman equation, but not
  explicitly modeled in the PC account that
  mostly emphasizes prediction about the
  immediate future.
  \item Mismatch negativity of PC is immanent in reward prediction error
  by the difference between actually predicted reward of an action given
  a state and the reward predicted by non-linear regression with graded
  discounting of future rewards.
  \item Sensory experience is a generative process from both views.
  In predictive coding, sensory perception of the external world
  is a generative experience due to modulatory top down experience at
  various hierarchical levels of sensory input processing.
  In our RL view, the (partially observed)
  environmental model is incorporated in the Markov decision
  process, \textit{which can be fully recovered based on the last
  state alone}.
  \item Adapting the neuronal connections for improved top-down modulation
  takes the concrete form of gradient computation
  \textit{and back-propagation} in MDPs and RL.
  \item Both provide a parsimoneous explanation why
  humans seem to minimize processing of incoming information
  when the environment is predictable and allocate attention when
  novel stimuli are encountered.
  Similarly, both propose explanations why
  predicting the future is inherently
  linked to information from the past.
\end{itemize}

\subsection{Semantic Hypothesis}
Another frequently proposed account of DMN function revolves
around forming meaning associations and analogies between
the currently experience and
abstract knowledge derived from past experiences
\cite{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge extractable from past memories
\cite{bar2009proactive}.
%
Going much beyond language ifself,
semantic analogies can be used to mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future analogical associations.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \cite{addis2008age, spreng2006temporal}.
A constructive account further explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
It is also closely related to episodic memory, language, problem solving,
planning, estimating others' thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\cite{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \cite{tolman1948cognitive}.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as the Markov property,
  that is, the current state directly results from the
  agent's history of states and actions. The learned
  value matrix and action transition schedules drive
  stimulus processing and action choice in the present.
  \item In both, semantic hypothesis and MDPs,
  there is no evidence to indicate that predictions of various
  levels of complexity, abstraction, timescale and purpose
  use mechanisms that are qualitatively different. This concurs with
  DMN activity increases across time, space, and content domains
  demonstrated in many neuroimaging studies. The semantic hypothesis
  and RL account provide explanations why hippocampus lesion does
  not only impair retrieving memories, but also hypothetical and future
  thinking \cite{hassabis2007patients}.
  \item Both the semantic hypothesis and MDP expose mechanisms of
  how alternative decision trees could be mentally explored.
  \item While mental scene construction implies a distinction between
  environmental perception and internally generated mind-wandering
  \footnote{``A computational model of how such a process might be structured
  is far from being defined, but it will probably require a form of
  regulation by which perception of the current world is suppressed
  while simulation of possible alternatives are constructed,
  followed by a return to perception of the present.''\cite{buckner2007self}},
  the MDP naturally integrates the former egocentric
  (related to current action, state and reward) and the later
  allocentric (related to past and future actions, states, and rewards)
  angles on the world in a same optimization problem.
  \item The semantic account of DMN function does not offer 
  a mechanistic explanation how semantic world knowledge and analogies thereof
  lead to prediction of future actions and states.
  \item In contrast to existing accounts on semantics and
  mental scene construction, the random and creative aspects of DMN function
  are here directly related to the advantages of stochastic optimization.
  \item The semantic hypothesis does not explain why memory recall
  for scene construction in humans is fragmentary and noisy
  instead of accurate and reliable.
  Yet, the MDP framework provides an algorithmic explanation in that
  stochasticity of the parameter space search implemented
  by Monte Carlo solvers provably yields better models of the world.
  That is, the purposeful stochasticity of policy and value estimation
  in MDPs provides a candidate explanation for why humans
  have evolved imperfect memories as the better adaptation.
  \item While both semantic and MDP account propose memory-based internally
  generated information for probabilistic mental models of action outcomes,
  only MDPs render explicit the grounds on which the final action is
  actually chosen (i.e., the estimated cumulative reward).
  \item In sum, episodic scene construction according to the semantic
  hypothesis lacks an explicit time and incentive model.
\end{itemize}

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constant, high energy costs?
MDPs provide an attractive
computational model how the human association cortex
might implement supra-modal control of the environment to
optimize the organism's fate by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN literature by a simple but non-trivial concept.
%
Viewed as a MDP, behavior unfolds by integrating across past
and future experience to guide action choice in the present context.
This account is more compatible with the DMN's involvement across
autobiographical memory retrieval, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
MDPs provide a mathematical formalism how
optimal substructure can be recursively exploited
when confronted with nested decisions.
Algorithmic gains by injecting stochasticity in past
behavior and outcomes may plausibilize why
very accurate memories have been evolutionarily de-selected
and why human creativity might be adaptive.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.
The proposed machine-learning
approach to DMN function is not only practically computable but
also neuroscientifically falsifiable.
At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies in DMN research.
%
Ultimately,
the DMN can be viewed as a smoothing kernel extending
from the relatively recent events to potentially upcoming events
to constantly improve sensory input and peformed action in the world.

>>>>

\section{Relation to other cognitive models}
\subsection{Predictive Coding Hypothesis}
The predictive coding and Bayesian brain framework 
is a frequently evoked idea related to the default mode function
\cite{bar2007}.
According to this functional account,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information. 
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
In fact,
back projections probably outnumber by far the existing input projection in the brain
\cite{salin1995corticocortical}.
This can explains the constructive, generative nature of sensory perception
\cite{friston2010free} and
why motor action is intimately related to sensory expectations
\cite{wolpert1995internal}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back-projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is maintained by top-down modulation of sensorimotor
processing by a-priori information about the environment.
The generative model of how data arise in the
world would hence be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the incoming sensory examples.

\begin{itemize}
  \item Both the predictive coding account and MPDs
  are process theories backed up by plausible
  neurobiological evidence that
  view the brain as a ``statistical organ''
  generalizing from the past to new experiences.
  \item Both expose a mechanisms of brain plasticity in that 
  neuronal wiring gets increasingly adapted
  when faced by unprecedented environmental
  challenges.
  \item The hierarchical aspects from predictive coding
  is re-expressed in MDPs in form of
  nested prediction of probable upcoming actions and rewards.
  \item Both model the consequences of action. In RL, the horizon of that
  look into the future is manifested in the $\gamma$ parameter
  in the Bellman equation, but not
  explicitly modeled in the PC account that
  mostly emphasizes prediction about the
  immediate future.
  \item Mismatch negativity of PC is immanent in reward prediction error
  by the difference between actually predicted reward of an action given
  a state and the reward predicted by non-linear regression with graded
  discounting of future rewards.
  \item Sensory experience is a generative process from both views.
  In predictive coding, sensory perception of the external world
  is a generative experience due to modulatory top down experience at
  various hierarchical levels of sensory input processing.
  In our RL view, the (partially observed)
  environmental model is incorporated in the Markov decision
  process, \textit{which can be fully recovered based on the last
  state alone}.
  \item Adapting the neuronal connections for improved top-down modulation
  takes the concrete form of gradient computation
  \textit{and back-propagation} in MDPs and RL,
  although this may not be neurobiologically plausible.
  \item Both provide a parsimonious explanation why the
  human brain decreases processing load devoted to incoming information
  when the environment is predictable and allocate increased
  attentional resources when novel stimuli are encountered.
  Similarly, both propose explanations why
  predicting the future is inherently
  linked to information from the past.
\end{itemize}

\subsection{Semantic Hypothesis}
Another frequently proposed account of DMN function revolves
around forming meaningful associations and analogies between
the current experience and
the abstract knowledge derived from past experiences
\cite{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge extractable from the environment
\cite{bar2009proactive}.
%
Going much beyond language ifself,
semantic analogies can be used to mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \cite{addis2008age, spreng2006temporal}.
A constructive account further explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
It is also closely related to episodic memory, language, problem solving,
planning, estimating others' thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\cite{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \cite{tolman1948cognitive}.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as the Markov property,
  that is, the current state directly results from the
  agent's history of states and actions. The learned
  value matrix and action transition schedules drive
  stimulus processing and action choice in the present.
  \item In both semantic hypothesis and MDPs,
  there is no evidence to indicate that predictions of various
  levels of complexity, abstraction, timescale and purpose
  use mechanisms that are qualitatively different. This concurs with
  DMN activity increases across time, space, and content domains
  demonstrated in many neuroimaging studies. The semantic hypothesis
  and RL account provide explanations why hippocampus lesion does
  not only impair retrieving memories, but also hypothetical and future
  thinking \cite{hassabis2007patients}.
  \item Both the semantic hypothesis and MDP expose mechanisms of
  how alternative decision trees could be mentally explored.
  \item While mental scene construction implies a distinction between
  environmental perception and internally generated mind-wandering\footnote{
  ``A computational model of how such a process might be structured
  is far from being defined, but it will probably require a form of
  regulation by which perception of the current world is suppressed
  while simulation of possible alternatives are constructed,
  followed by a return to perception of the present.''\cite{buckner2007self}},
  the MDP naturally integrates the former egocentric
  (related to current action, state and reward) and the later
  allocentric (related to past and future actions, states, and rewards)
  angles on the world in a same optimization problem.
  \item The semantic account of DMN function does not offer 
  a mechanistic explanation how semantic world knowledge and analogies thereof
  lead to prediction of future actions and states.
  \item In contrast to existing accounts on semantics and
  mental scene construction, the random and creative aspects of DMN function
  are explained in MDPs by the advantages of stochastic optimization.
  \item The semantic hypothesis does not explain why memory recall
  for scene construction in humans is typically fragmentary and noisy
  instead of accurate and reliable.
  Yet, the MDP framework provides an algorithmic explanation in that
  stochasticity of the parameter space search implemented
  by Monte Carlo solvers provably yields better models of the world.
  That is, the purposeful stochasticity of policy and value estimation
  in MDPs provides a candidate explanation for why humans
  have evolved imperfect memories as the better adaptation.
  \item While both semantic and MDP account propose memory-based internally
  generated information for probabilistic mental models of action outcomes,
  only MDPs render explicit the grounds on which the final action is
  eventually chosen (i.e., the estimated cumulative reward).
  \item In sum, episodic scene construction according to the semantic
  hypothesis is lacking an explicit time and incentive model.
\end{itemize}

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constant, high energy costs?
MDPs provide an attractive
computational model how the human association cortex
might implement supra-modal control of the environment by
obtaining reward and avoiding pain to
optimize the organism's fate.
This idealized process model explains
a number of previous experimental observations in the
DMN literature by a simple but non-trivial concept.
%
Viewed as an MDP, behavior unfolds by integrating happened past events
and possible future events to guide action choice in the present context.
This functional account is more compatible with the DMN's
poorly understood involvement across
autobiographical memory retrieval, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
MDPs provide a mathematical formalism how
optimal substructure can be recursively exploited
when confronted with nested decisions.
Algorithmic gains by injecting stochasticity in the recall of past
behavior and outcomes may plausibilize why
very accurate memories have been evolutionarily disfavored
and why human creativity might be adaptive.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.
The proposed machine-learning
approach to DMN function is hence not only practically computable but
also neuroscientifically falsifiable.
At the least, we propose an alternative vocabulary to
describe and interpret neuroscientific studies in DMN research.
%
Ultimately,
the DMN can be viewed as a smoothing kernel extending
from the relatively recent events to potentially upcoming events
to constantly improve sensory processing and
action performance in the world.
